Predicting the Ending Notes of Bach Chorales

### 2018-04-26 Paper.tex
Finish first draft of essay

\section{Data}
The data set used to test the models is a collection of one hundred Bach chorales, provided by the University of California Irvine's Machine Learning Repository \cite{data_set}. Each chorale contains only the melody line from each chorale. Each note in the melody has six different information fields. The starting time of the note is measured in 16th notes from the start of the piece. The pitch of the note is provided in the MIDI tuning standard, with the note C4 given a value of 60, C\#4 given a value of 61, etc. The pitches are restricted to values between 60 and 75, by nature of the notes present in the chorales. The duration of each note is given, also measured in 16th notes. The key signature is presented as values between -4 and +4. The numbering relates to the number of sharps or flats are present in the key signature: positive for sharps, negative for flats. The time signature is again measured in the number of 16th notes, this time in the number of 16th notes per measure. Finally, a Boolean flag is set to true if the note is underneath a fermata.

@@

\begin{figure}[h]
  \includegraphics[width=13.5cm]{Naive}
  \centering
  \caption{A confusion matrix of the naive model. The green region represents overestimates, the blue region represents underestimates.}
\end{figure}

@@

@@

This model predicted the endings of a total of 56 chorales correctly. Despite only a minor increase in correct pitches, the MSE decreased dramatically to 4.32. While the number of underestimations was reduced by around 60\%, there was a 50\% rise in the number of overestimations. This switch may be caused by a number of possibilities. Since the model only selected notes within the major triad, any chorale which ended on a minor third would be missed by the model.

\subsection{Computer Model}
In the final model, each chorale was used in a decision tree to predict the final pitch. 80 random chorales from the data set were used as the training set, with the remaining 20 chorales left for testing. Five total trees were created, with a different subset of 20 each time; these five trees therby predicted over the entire data set.\\

This model was able to correctly predict 55 of the 100 chorales. 21 chorales were overestimated, and 24 chorales were underestimated, with a MSE of 15.93. Looking at the various decision trees generated by the model, most trees contained the same node: checking whether the starting key was either A or E major. In some cases, there would be only one solution for chorales which were in A and E major, usually MIDI note 69 (A4). Overall, the most prevalent node checked the starking key signature. After key signature, the most prevalent nodes checked the pitch of the first note. While these covered the majority of the nodes within each tree, there were a number of outlier nodes. For instance, one fairly common outlier was built on the pitch of the 16th or 17th note in the melody line; this note is likely the end of the first phrase of the melody. Other nodes were related to the overall speed of the chorale; one node checked the length of the 15th note, and another checked if the 34th note of the melody was on the 34th beat or later. There may be a pattern apparent in the data, where faster chorales are more likely to end on a particular note than slower chorales.\\

Although the human and computer models performed at similar levels, one noticeable difference between the two is accuracy. The confusion matrix for the computer model had fewer rows and columns than the human model. The pitches for the notes were given in a range between 60 and 75 in the MIDI scale, but this range applies to the entire melody line, not the endings. While the human model did not assume any distribution of pitches in the range, the computer model determined that the ending notes are clustered in the middle of the range, thus explaining the shrinkage of the confusion matrix.\\

@@

@@

In an attempt to prune satistically insignificant data, the tree model was run again with the starting time, length, and time signature removed. This variation predicted 54 chorale endings correctly, only one less than the original computer model, and had a MSE of 14.94. However, the split between underestimates and overestimates was worse; 30 chorales were underestimated and 16 were overestimated. Another variation was created with the same fields removed as above, but with an extra caveat that only notes under a fermata were used to predict the endings. This variation was still able to predict 50 chorales correctly, with a similar spread of over- and underestimates as the original computer model: 28 under, 22 over, MSE of 17.36. Clearly, the most useful predictors are found at the ends of musical phrases, but there are still insights to be gained from the other notes.
\section{Conclusion}
Although the computer model performed at a similar level to the human model in total correct, the human model had greater precision on its incorrect guesses than any variation of the computer model. The classification trees that were generated had less to do with actual patterns of notes in the music, and had more in common with the naive model than the human model. Ideally, a computer that truly thought like a human would, in this scenario, look at more than just key signature.\\

>>>>

\section{Data}
The data set used to test the models is a collection of one hundred Bach chorales, provided by the University of California Irvine's Machine Learning Repository \cite{data_set}. Each chorale contains only the melody line from each chorale. Each note in the melody has six different information fields. The starting time of the note is measured in 16th notes from the start of the piece. The pitch of the note is provided in the MIDI tuning standard, with the note C4 given a value of 60, C\#4 given a value of 61, etc. The pitches are restricted to values between 60 and 75, by nature of the notes present in the chorales. The duration of each note is given, also measured in 16th notes. The key signature is presented as values between -4 and +4. The numbering relates to the number of sharps or flats are present in the key signature: positive for sharps, negative for flats. The time signature is again measured in the number of 16th notes, this time in the number of 16th notes per measure. Finally, a Boolean flag is set to true if the note is underneath a fermata. Fermatas generally indicate the end of a musical phrase or chord progression, and instruct the performer to hold the note for as long as deemed necessary.

@@

\begin{figure}[h]
  \includegraphics[width=13.5cm]{Naive}
  \centering
  \caption{A confusion matrix of the naive model. The indices relate to the MIDI values of the predicted/actual notes. The green region represents overestimates, the blue region represents underestimates.}
\end{figure}

@@

\begin{figure}[h]
  \includegraphics[width=13.5cm]{Human}
  \centering
  \caption{A confusion matrix of the human model. As in Figure 1, indices are MIDI pitches, the green region is overestimates, and the blue region is underestimates.}
\end{figure}

@@

This model predicted the endings of a total of 56 chorales correctly. Despite only a minor increase in correct pitches, the MSE decreased dramatically to 4.32. While the number of underestimations was reduced by around 60\%, there was a 50\% rise in the number of overestimations. This switch may be caused by a number of possibilities. Since the model only selected notes within the major triad, any chorale which ended on a minor third would be missed by the model. There are likely some patterns which are not considered by the algorithm, where the two penultimate notes have two possible endings. In these cases, only the more prevalent one was incorporated into the algorithm. Finally, any chorale which uses the last two notes as part of a key change will not be accurately predicted, since this model uses the relative position within the current key.

\subsection{Computer Model}
In the final model, each chorale was used in a decision tree to predict the final pitch. 80 random chorales from the data set were used as the training set, with the remaining 20 chorales left for testing. Five total trees were created, with a different subset of 20 each time; these five trees thereby predicted over the entire data set.\\

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=7cm]{Tree1}
  \includegraphics[width=7cm]{Tree2}\\
  \centering
  \includegraphics[width=7cm]{Tree3}
  \includegraphics[width=7cm]{Tree4}
  \caption{The four other trees generated by the computer model. }
\end{figure}

This model was able to correctly predict 55 of the 100 chorales. 21 chorales were overestimated, and 24 chorales were underestimated, with a MSE of 15.93. Looking at the various decision trees generated by the model, most trees contained the same node: checking whether the starting key was either A or E major. In some cases, there would be only one solution for chorales which were in A and E major, usually MIDI note 69 (A4). Overall, the most prevalent node checked the starting key signature. After key signature, the most prevalent nodes checked the pitch of the first note. While these covered the majority of the nodes within each tree, there were a number of outlier nodes. For instance, one fairly common outlier was built on the pitch of the 16th or 17th note in the melody line; this note is likely the end of the first phrase of the melody. Other nodes were related to the overall speed of the chorale; one node checked the length of the 15th note, and another checked if the 34th note of the melody was on the 34th beat or later. There may be a pattern apparent in the data, where faster chorales are more likely to end on a particular note than slower chorales.\\

Although the human and computer models performed at similar levels, one noticeable difference between the two is accuracy. The confusion matrix for the computer model has one less row and column than both the naive model and 3 fewer rows and columns than the human model. The pitches for the notes were given in a range between 60 and 75 in the MIDI scale, but this range applies to the entire melody line, not the endings. While the human model did not assume any distribution of pitches in the range, the computer model determined that the ending notes are clustered in the middle of the range, thus explaining the shrinkage of the confusion matrix.\\

\begin{figure}[h]
  \includegraphics[width=13.5cm]{confusionMat}
  \centering
  \caption{The confusion matrix generated by the computer model. As in Figure 1, indices are MIDI pitches, the green region is overestimates, and the blue region is underestimates. In comparison with the matrix in Figure 1, this matrix has one fewer row and column.}
\end{figure}

@@

In an attempt to prune statistically insignificant data, the tree model was run again with the starting time, length, and time signature removed. This variation predicted 54 chorale endings correctly, only one less than the original computer model, and had a MSE of 14.94. However, the split between underestimates and overestimates was worse; 30 chorales were underestimated and 16 were overestimated. Another variation was created with the same fields removed as above, but with an extra caveat that only notes under a fermata were used to predict the endings. This variation was still able to predict 50 chorales correctly, with a similar spread of over- and underestimates as the original computer model: 28 under, 22 over, MSE of 17.36. Clearly, the most useful predictors are found at the ends of musical phrases, but there are still insights to be gained from the other notes.
\section{Conclusion}
Although the computer model performed at a similar level to the human model in total correct, the human model had greater precision on its incorrect guesses than any variation of the computer model. The classification trees that were generated had less to do with actual patterns of notes in the music, and had more in common with the naive model than the human model. Ideally, a computer that truly thought like a human would, in this scenario, look at more than just key signature and the first pitch.\\
